Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-07-03 hazelnut-popl17.tex
added colored highlighting to zipper focus; adding sum types

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
 Calculus}
%\subtitle{Modular Type Constructors}

@@

%  \emph{Structure editors} have long promised to alleviate these burdens by exposing only edit actions that  cause sensible changes to the program structure.
% Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}. They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.
% Structure editors (e.g. Scratch) eliminate the possibility of syntax errors.
\todo{why not title just ``hazelnut: a structure editor calculus''?}
\emph{Structure editors} allow programmers to edit the tree structure of a program directly -- there are no syntax errors. This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they need not contend with malformed program text. %Structure editors for professional programmers are also under active development (e.g. \texttt{mbeddr}.)

This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes this approach two steps further: Hazelnut's edit actions operate over terms that are not just syntactically well-formed but also statically and dynamically meaningful.
Na\"ively, this injunction against type errors would force programmers to construct terms in an unnatural ``outside-in'' manner. To avoid this problem, Hazelnut's {action semantics} automatically places terms that have a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in gradual type systems.

We intend Hazelnut as a minimal foundational calculus, rather than as an end-user tool itself. To that end, we show how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to Hazelnut. We also briefly discuss how Hazelnut's semantics lends itself to implementation as a functional reactive web application. Our reference implementation is written using \lstinline{js_of_ocaml}.

@@

>>>>

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
 Calculus}

@@

%  \emph{Structure editors} have long promised to alleviate these burdens by exposing only edit actions that  cause sensible changes to the program structure.
% Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}. They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.
% Structure editors (e.g. Scratch) eliminate the possibility of syntax errors.

\emph{Structure editors} allow programmers to edit the tree structure of a program directly. This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they do not need to contend with malformed program text.

This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{focus} (\emph{ala} Huet's zipper.) Hazelnut goes one step beyond syntactic well-formedness: it's {edit actions} operate over statically meaningful terms.  
Na\"ively, this prohibition on ill-typed edit states would force the programmer to construct terms in a rigid ``outside-in'' manner. To avoid this problem, the {action semantics} automatically places terms assigned a type that is inconsistent with the expected type {inside} a {hole}. This safely defers the type consistency check until the term inside the hole is \emph{finished}.

Hazelnut is a foundational type-theoretic account of typed structure editing, rather than an end-user tool itself. To that end, we describe how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to the calculus. We also discuss various plausible evaluation strategies for terms with holes, and in so doing reveal connections with gradual typing and contextual modal type theory (the Curry-Howard interpretation of contextual modal logic.) Finally, we  discuss how Hazelnut's semantics lends itself to implementation as a functional reactive program. Our reference implementation is written using \lstinline{js_of_ocaml}.

@@

In addition to eliminating malformed edit states, structure editors also promise to increase the speed of development. However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g. whether a keyboard or a mouse is in use.\todo{integrate this}

By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.  
By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?  In progress, unfinished, incomple?} and yet still in a well defined type state. 
This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.\todo{integrate this}

### 2016-07-03 intro.tex

%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
% Programming languages, and therefore programs themselves, are rich mathematical structures. 
Well-typed programs are rich semantic structures, but human programmers typically construct and manipulate them only indirectly: by editing text that is parsed according to a textual syntax and then typechecked according to a static semantics. This indirection is of substantial practical utility, to be sure -- text editors and other text-based tools benefit from decades of development effort. However, it also introduces some fundamental complexity into the programming process.

@@

Moreover, the fact that not all sequences of characters correspond to a well-formed, meaningful program complicates matters for tool designers. Tools are confronted with such text quite frequently -- by analyzing a dataset gathered by Yoon and Myers \cite{6883030} (which contains 1,460 hours of fine-grained edit logs collected from 21 programmers) we found that 44.2\% of code edits resulted in malformed edit states. Some additional percentage of edit states were well-formed but ill-typed (the dataset did not include enough information to determine the exact percentage.)  
Some of these edit states were due to programmer error, while others arose simply because the programmer was in the midst of a sequence of edit actions that left the text temporarily malformed or ill-typed. The language definition is silent about these edit states, so it is difficult to design useful editor services (e.g. syntax highlighting~\cite{sarkar2015impact} and semantics-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348}) and report meaningful error messages\todo{citations}.
%, or because the programmer has made a mistake. 
% The syntax definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action

Some editors have developed \emph{ad hoc} workarounds for this problem, e.g. they might attempt to insert closing delimiters automatically, use whitespace to guess where a construct is likely to end, use regular expressions to highlight malformed program text, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are imperfect, complex and sometimes misleading. % This may also help explain why the error messages emitted by parsers and typecheckers are often quite baroque\todo{citations}.

@@

% The type structure of the language makes this sort of error obvious: it's
% not that you're adding characters that make your program incorrect, or even
% malformed; you're adding characters that can't possibly create a structure
% you want because of the type. Simply put, the primitive operations
% available in text editors do not always correspond to sensible
% transformations on the structure of the program.
One attempt to solve this problem has been \emph{Structure editors},
where every edit action leaves the program being manipulated in a 
structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also often promised that programs can be written more quickly using a structure editor. However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g. whether a keyboard or a mouse is in use.} 
%Scratch etc. for novices. mbeddr for pros. Lamdu and Unison ongoing efforts with a functional foundation. We give other examples of notable structure editors in Sec. \ref{sec:rw}.\todo{revise}

However, these syntactically correct states can be semantically meaningless (i.e. undefined), because the language definitions generally only give meaning to complete, well-typed terms. (e.g., when a branching element is introduced, the program does not become semantically correct until both branches are complete)
This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor.
Similar to text editors, structure editors also can develop workarounds to try to help users, but these efforts can only extend so far, because of the underlying language definition.
%Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g. Unison ; Scratch seems not to allow literals of the wrong type, but variables are not typed.)\todo{revise} %More sophisticated semantic reasoning principles thus remain .

%% TODO: some sort of transition that talks about how you still don't have
%% semantic reasoning principles because you can be left in a semantically
%% ill-defined state

%\todo{revise remainder of intro}

In this paper, we present a minimal structure editor, Hazelnut,  defined in the type-theoretic style (i.e. Hazelnut is a \emph{structure editor calculus}). 
By defining \emph{holes} as a language construct, for both expressions as types, Hazelnut enables type-aware actions that always leave the program in both a structurally and semantically well-defined state. 
%In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics. Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state. 
In fact, edit actions maintain an even stronger invariant -- when acting on an expression whose type is determined by its surroundings (e.g. an expression in function argument position), only edit actions consistent with that type are permitted (we will formally state this invariant in Sec. \ref{sec:actions}.) This does not imply that programs need to be constructed in a strictly outside-in manner, however, because an expression that has a type that is not yet consistent with its surroundings will automatically be put into a {hole} that defers consistency analysis until the hole is \emph{finished}.

By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.  
By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?  In progress, unfinished, incomple?} and yet still in a well defined type state. 
This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.

\todo{Talk about dynamics.}

The remainder of the paper is organized as follows:
\begin{itemize}
  \item We begin with an example to develop the reader's intuitions in Section
    \ref{sec:example}.

  \item We then formally define Hazelnut in Section \ref{sec:hazel} and state the important metatheoretic properties.
  \item Dynamics \ref{sec:dynamics}

  \item In Section \ref{sec:mech}, we describe our ongoing effort to formalize the semantics and metatheory of Hazelnut in Agda.

  \item In Section \ref{sec:impl}, we describe our ongoing effort to implement Hazelnut in a web browser, using a functional reactive model for user interaction.
  \item In Section \ref{sec:rw}, we give an overview of related work.
  \item We conclude in Section \ref{sec:future} by describing our vision for this work going 
    forward.
\end{itemize}

>>>>

%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
% Programming languages, and therefore programs themselves, are rich mathematical structures. 
Well-typed programs are rich semantic structures, but human programmers typically construct and manipulate them only indirectly: by editing text that is parsed according to a textual syntax and then typechecked according to a static semantics. This indirection is of substantial practical utility, to be sure -- text editors and other text-based tools benefit from decades of development effort -- but it also introduces some fundamental complexity into the programming process.

@@

The fact that not every sequence of characters corresponds to a meaningful program also complicates matters for tool designers. In particular, program editors confront meaningless text quite frequently: in a dataset collected by Yoon and Myers consisting of 1460 hours of fine-grained edit logs from 21 programmers, 44.2\% of edits resulted in a malformed edit state \cite{6883030}. Some additional percentage of edit states were well-formed but ill-typed (the dataset did not record the imported libraries, so an exact percentage could not be determined.)  
Some of these edit states could be attributed to programmer error, while others arose simply because the programmer was in the midst of a sequence of edit actions that left the text temporarily malformed or ill-typed. The language definition is silent about these edit states, so it is difficult to design useful editor services (e.g. syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring \todo{citation}) and to report localized, meaningful error messages\todo{citations}.
%, or because the programmer has made a mistake. 
% The syntax definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action

Some editors have developed \emph{ad hoc} workarounds for this problem, e.g. they might attempt to use regular expressions to highlight malformed program text, insert closing delimiters automatically, use whitespace to guess where a delimiter is likely to appear, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.) These heuristic methods are often complex, and they can confuse or mislead the programmer. % This may also help explain why the error messages emitted by parsers and typecheckers are often quite baroque\todo{citations}.

@@

% The type structure of the language makes this sort of error obvious: it's
% not that you're adding characters that make your program incorrect, or even
% malformed; you're adding characters that can't possibly create a structure
% you want because of the type. Simply put, the primitive operations
% available in text editors do not always correspond to sensible
% transformations on the structure of the program.
These complications have motivated a long line of research into \emph{structure editors}, i.e. program editors where edit actions maintain non-trivial structural invariants.

Most structure editors are \emph{syntactic structure editors}, i.e. the edit states are trees with \emph{holes} representing branches that have yet to be constructed, and the edit actions are context-free tree transformations. Syntactic structure editors like Scratch \cite{Resnick:2009:SP:1592761.1592779} are popular in childhood educational settings, in part because they eliminate the possibility of syntax errors\todo{papers with case studies?}.

The Scratch language has a trivial static semantics (i.e. it requires only that every assignable has been declared) and an \emph{ad hoc} dynamic semantics (i.e. if evaluation encounters a hole or a dynamic error, it goes on to the next command.) Researchers have also developed syntactic structure editors for  languages with a more conventional semantics. For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}. These editors all have interesting user interfaces, but there are several foundational problems that complicate their design. 
% \begin{itemize}
% \item

First, there is an all too familiar problem: these languages do not have a clear formal semantics and metatheory, i.e. ``the implementation is the specification''. This makes it difficult to reason methodically about types and binding, and therefore, to design sound and useful editor services.

Developing a formal semantics and metatheory in the type theoretic style for the underlying language (e.g. following Standard ML \cite{Harper00atype-theoretic,mthm97-for-dart}) would go only part of the way towards addressing the challenges faced by structure editor designers, because a standard semantics assigns no formal meaning to \emph{incomplete terms}, i.e. terms with holes. Moreover, a syntactic structure editor does not guarantee that every edit state is statically meaningful, so incorporating holes into the static semantics would in any case be only a minor salve. Heuristics analagous to those that pervade textual program editors would still be needed to contend with the ill-typed edit states.

%This brings us to the contributions of this paper. 
This paper defines Hazelnut,  a \emph{typed structure editor calculus} that addresses these foundational problems. 
In particular, Hazelnut is organized around a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}. Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e. well-typed) expression with a single \emph{focus} of action (following Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.) Actions act at the expression or type in focus (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g. a function argument), only actions consistent with that type are permitted.

Na\"ively, enforcing this injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner. For example, the programmer would often need to construct the outer ``function application'' form before identifying the intended function. To avoid this problem, the action semantics automatically places expressions assigned a type that is inconsistent with the expected type inside a {hole}. This safely defers the type consistency check until the expression inside the hole is \emph{finished}. In other words, holes appear both at the leaves of the tree and at the internal nodes that remain under construction. %In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}. %Actions act at a programmer-indicate subtree, called the \emph{focus} (which is defined following Huet's zipper pattern.)

% By defining \emph{holes} as a language construct, for both expressions as types, Hazelnut enables type-aware actions that always leave the program in both a structurally and semantically well-defined state. 
%In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics. Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state.

%However, these syntactically correct states can be semantically meaningless (i.e. undefined), because the language definitions generally only give meaning to complete, well-typed terms. (e.g., when a branching element is introduced, the program does not become semantically correct until both branches are complete)
% This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor.
% Similar to text editors, structure editors also can develop workarounds to try to help users, but these efforts can only extend so far, because of the underlying language definition.
%Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g. Unison ; Scratch seems not to allow literals of the wrong type, but variables are not typed.)\todo{revise} %More sophisticated semantic reasoning principles thus remain .

The remainder of the paper is organized as follows. We begin with two example edit sequences to develop the reader's intuitions in Section
    \ref{sec:example}. We then give a detailed overview of Hazelnut's semantics and metatheory, which has been mechanized in Agda, in Section \ref{sec:hazel}. Hazelnut is designed as a {foundational} calculus, i.e. as a calculus that language and editor designers extend with higher level constructs. We show how Hazelnut's rich mechanized metatheory guides and constrains the development of such an extension in Section \ref{sec:extending}.  We consider several possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}. In so doing, we discover interesting connections with gradual typing and contextual modal type theory. The former provides an interpretation of type holes, and the latter provides a logical interpretation of expression holes. It also suggests a principled design for an ``edit and resume'' feature. In Section \ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program. Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library. In Section \ref{sec:rw}, we summarize related work. We conclude in Section \ref{sec:future} by describing our vision of a principled science of structure editing.
\end{itemize}
